---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# dominoise

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![R-CMD-check](https://github.com/InseeFrLab/dominoise/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/InseeFrLab/dominoise/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Protect magnitude tables — turnover, payroll, output — against disclosure, and
know exactly what that protection costs. `dominoise` implements a perturbation
mechanism whose risk and utility metrics are available in **closed form**, so
that choosing its parameters is an analytical decision rather than an empirical
one: no simulation, no recalibration on the data.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("InseeFrLab/dominoise")
```

## The mechanism

Each positive total `Y` is multiplied by a Gaussian noise combining two
components:

$$Y' = Y\,(1 + \rho^{\,n}\nu + \varepsilon), \qquad
\nu \sim \mathcal{N}(0,\sigma_\nu^2),\quad
\varepsilon \sim \mathcal{N}(0,\sigma_\varepsilon^2)$$

where `rho = X1 / Y` is the share of the largest contributor in the total.

The **dominance-driven** component `rho^n * nu` concentrates the perturbation on
the cells that a dominant contributor makes sensitive; its weight fades as `rho`
moves away from 1, the faster the larger `n`. The **general-purpose** component
`eps` perturbs every cell at a uniform relative rate and carries the residual
protection against differencing, where the risk is not borne by a dominant
contributor. Three parameters govern the whole mechanism: `sigma_nu`,
`sigma_eps` and `n`.

From this distribution, the package derives in closed form the information loss
and a risk measure for each of three attack scenarios:

| Scenario | Attacker | Rule transposed |
|---|---|---|
| **I** — external inference | attributes the whole published total to the dominant contributor | dominance rule (1, k) |
| **II** — internal inference | a contributor subtracts their own value from the total | p%-rule |
| **DIFF** — differencing | subtracts two totals known to differ by a single unit | — |

## Expected data

`dominoise` works on an **already aggregated** table, with **already aggregated**
cell keys:

| Column | Required | Content |
|---|---|---|
| `total` | yes | the aggregate `Y`, strictly positive |
| `x1` | yes | largest individual contribution, same unit as `total` |
| `x2` | no | second largest contribution — enables the scenario-II assessment |
| `ck_var` | yes | cell key, in `[0;1]` (or set `reduce_key = TRUE` to take the fractional part of a raw sum of individual keys) |

```{r data}
library(dominoise)

set.seed(4081789)
n_cells <- 500
turnover <- round(runif(n_cells, 1e3, 1e6))
share    <- rbeta(n_cells, 2, 2)
tab <- data.frame(
  turnover = turnover,
  x1       = round(share * turnover),
  x2       = round(0.5 * (1 - share) * turnover),
  ck       = runif(n_cells)
)
head(tab, 3)
```

## The workflow

Calibration proceeds in two steps, each following the same pattern: **explore**
a theoretical grid, **look** at the trade-off, **commit** the decision into the
parameter object. Everything is data-free until the last step.

### 1. The differencing noise sets `sigma_eps`

`sigma_eps` is the only parameter able to control the differencing risk, and
also the one that sets the loss floor borne by *every* cell, including the least
sensitive. Choose an accuracy threshold `beta` and a risk ceiling `tau`; the
required `sigma_eps` follows analytically.

```{r step1}
grid_diff <- pm_calib_diff(tau = seq(0.80, 0.95, 0.01))
head(grid_diff)

params <- pm_params()
params <- pm_commit_diff(params, beta = 0.05, tau = 0.90)
```

### 2. The dominance noise sets `sigma_nu` and `n`

At that fixed `sigma_eps`, the scenario-I risk and the information loss both
depend on `sigma_nu` and `n`. The two objectives move in opposite directions
with `n`: utility improves at every `rho < 1`, while the worst-case risk rises.
The rule is therefore to keep the **largest** `n` that still meets the ceiling.

```{r step2}
grid_dom <- pm_calib_dominance(params, sigma_nu = c(0.3, 0.4, 0.5))
summary(grid_dom)

pm_suggest_n(params, sigma_nu = c(0.3, 0.4, 0.5))
```

The trade-off map traces, for each candidate pair, the risk against the
information loss as `rho` runs over `]0;1]` — the producer reads off both the
worst-case risk and what it costs in utility:

```{r tradeoff, fig.height = 3.5, fig.width = 9}
plot(grid_dom)
```

```{r step2-commit}
params <- pm_commit_dominance(params, sigma_nu = 0.4, n = 4)
params
```

### 3. Apply the noise

```{r apply}
res <- pm_perturb(
  tab,
  total  = "turnover",
  x1     = "x1",
  x2     = "x2",
  ck_var = "ck",
  params = params
)
head(res[, c("turnover", "x1", "rho", "ck_nu", "turnover_pert")], 3)
```

### 4. Check what was achieved

Because the mechanism is analytical, the loss and the attack success rate
observed on the table should track the values predicted before any data were
touched. That agreement is the natural consistency check to present to a
disclosure review board.

```{r assess}
assess_utility_empirical(res)
assess_risk_empirical(res, scenario = "I")
```

## Function reference

Two prefixes carry the whole API. `pm_` for the functions that calibrate the
parameters or apply the noise; `assess_` for the ones that measure risk and
utility.

| Function | Purpose |
|---|---|
| `pm_params()` | container for every policy and mechanism choice |
| `pm_calib_diff()`, `pm_commit_diff()` | step 1 — explore and set `sigma_eps` |
| `pm_calib_dominance()`, `pm_suggest_n()`, `pm_commit_dominance()` | step 2 — explore and set `sigma_nu` and `n` |
| `pm_plot_risk_profile()`, `pm_plot_risk_max()`, `pm_plot_tradeoff()` | the paper's calibration figures |
| `pm_draws()`, `pm_perturb()` | derive the draws from the keys, apply the mechanism |
| `assess_loss_expectation()`, `assess_loss_ci()` | a-priori utility, conditional on `rho` |
| `assess_risk_I()`, `assess_risk_II()` | a-priori risk measures |
| `assess_utility_empirical()`, `assess_risk_empirical()` | ex-post metrics on the perturbed table |

## Limitations

**Scenario II is not controlled in its worst case.** Two informed contributors
sharing the whole of a cell cannot be protected at perturbation levels
compatible with acceptable utility. As soon as the configuration is slightly
relaxed — the two largest contributors covering, say, 95% of the total rather
than 100% — the mechanism does provide meaningful protection.

**The noise is multiplicative**, hence ineffective on zeros and values close to
zero. `pm_perturb()` leaves non-positive totals untouched and warns about them;
they need a separate treatment.

**Risk measures are conditional** on the attack scenario occurring, and on the
actual dominance of the cell, which only the producer knows.

## Governing the keys

Consistency over time — the same cell always receiving the same perturbation —
rests on the reproducibility of the hashed string, which combines the cell key,
the **indicator name** and the **aggregation operation**. Two consequences to
plan for before production:

- renaming an indicator between two releases changes every draw for that
  indicator, and reopens the averaging attack the keys are meant to prevent;
- `key_digits` (9 by default) fixes how the numeric key is turned into text
  before hashing. Changing it changes every draw. It is recorded in the
  `pm_meta` attribute of the perturbed table — archive it with the output.

## Reference

Jamme, J. (2026). *Parsimonious perturbation mechanism for magnitude tables with
analytical risk-utility metrics.*
[ResearchGate](https://www.researchgate.net/publication/408487480_Parsimonious_perturbation_mechanism_for_magnitude_tables_with_analytical_risk-utility_metrics)

All the results of the paper are reproducible from a companion repository:
[InseeFrLab/perturbation-mechanism-for-aggregates](https://github.com/InseeFrLab/perturbation-mechanism-for-aggregates).

```{r citation, eval = FALSE}
citation("dominoise")
```

## License

GPL-3.
